---
title: "Smoking, Age, and Mortality: A Longitudinal Analysis of Whickham Data"
format: 
  html:
    embed-resources: true
---

### Packages

We'll use the **tidyverse** package for much of the data wrangling and visualization and the data lives in the **mosaicData** package.


install.packages("mosaicData")

You can load them by running the following in your Console:

```{r}
library(tidyverse) 
library(mosaicData) 
```



## Solutions

::: {#sol-1}

This is an observational study, because the researchers recorded existing behaviors (like smoking) and outcomes (alive/dead) without assigning treatments or interventions. No random assignment occurred.

:::  



::: {#sol-2}

```{r}
nrow(Whickham)
```

The dataset contains 1314 rows, and each observation represents one individual in the study, recording their smoking status, age and outcome (alive or dead).


:::



::: {#sol-3}

The dataset contains three variables: age (numerical), smoker (categorical nominal), and outcome (categorical nominal).

```{r}
#| label: hist-bar-uni

# age: 
ggplot(Whickham, aes(x = age)) +
  geom_density(color = "blue", linewidth = 0.4, fill = "lightblue", alpha = 0.4) +
  labs(
    title = "Age Distribution of Participants",
    x = "Age",
    y = "Density"
  ) +
  theme_minimal()

# smoker: 
ggplot(Whickham, aes(x = smoker, fill = smoker)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Smoking Status Distribution",
    x = "Smoker",
    y = "Count"
  ) +
  theme_minimal() 


# outcome:  
ggplot(Whickham, aes(x = outcome, fill = outcome)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Health Outcome After 20 Years",
    x = "Outcome",
    y = "Count",
    fill = "Health Outcome"
  ) +
  theme_minimal() 

```

:::  




::: {#sol-4}

The expectation would be that smokers will have a higher proportion of deaths compared to non-smokers, since smoking is a known health risk.

:::




::: {#sol-5}

We create a Filled Bar Chart.

```{r}
#| label: bar-filled
Whickham |> 
  ggplot(aes(x = smoker, fill = outcome)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Proportion of Health Outcomes by Smoking Status",
    x = "Smoking Status",
    y = "Proportion",
    fill = "Health Outcome"
  ) +
  theme_minimal()

```

Description: From observing the proportion of outcomes within each smoking category, the data shows that smokers have a lower proportion of deaths and a higher proportion of survival compared to non-smokers.


We compute the conditional probabilities, which further confirm that smokers have a lower proportion of deaths and a higher proportion of survival (around 8%), compared to non-smokers—contrasting with our initial expectation.

```{r}
Whickham |> 
  select(smoker, outcome) |> 
  table() |>
  prop.table(margin = 1)
```


:::  




::: {#sol-6}

We need to introduce a new variable to Whickham, so we will need to use `mutate()`. 

```{r}
Whickham <- Whickham |> 
  mutate(age_cat = case_when(
    age <= 44 ~ "18-44",
    age > 44 & age <= 64 ~ "45-64",
    age > 64 ~ "65+"
  ))

head(Whickham)

```

:::


::: {#sol-7}


```{r}
#| label: bar-filled-byage
Whickham |> 
  ggplot(aes(x = smoker, fill = outcome)) +
  geom_bar(position = "fill") +
  facet_wrap(~ age_cat) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Health Outcome by Smoking Status & Age",
    x = "Smoking Status",
    y = "Proportion",
    fill = "Health Outcome"
  ) +
  theme_minimal()
  
```

Faceting the plot by age_cat changes our initial interpretation: within each age group, smokers tend to have a higher proportion of deaths and a lower proportion of survival. The faceting helps reveal patterns that were previously masked by age differences in the overall population.

:::  



::: {#sol-8}


```{r}
Whickham |>
  count(age_cat, smoker, outcome) |>
  group_by(age_cat, smoker) |>
  mutate(prop = n / sum(n))
```

These conditional proportions confirm the trend revealed by faceting: within each age group, smokers have a higher proportion of deaths and a lower proportion of survival than non-smokers. This contradicts the misleading initial interpretation from the aggregated data, which is mainly influenced by the greater number of older individuals among non-smokers.

:::




